---
title: "Data Analysis with R"
author: "UCLA Stats Club / DataFest Planning Committee"
date: "2/6/2019"
output:
  pdf_document:
    latex_engine: xelatex
mainfont: Open Sans
monofont: Courier
---

# Installing and loading the Tidyverse

```{r, message = FALSE, tidy.opts=list(width.cutoff=60)}
# install.packages("tidyverse")
library(tidyverse)
```

# Applying the Tidyverse: the 2017 grades dataset

```{r, message = FALSE, tidy.opts=list(width.cutoff=60)}
grades_2017 <- read_csv("grades_2017.csv")
names(grades_2017)
```

We can use dplyr's 'mutate' to create a variable, 'total_students':

```{r, message = FALSE, tidy.opts=list(width.cutoff=60)}
grades_2017 <- grades_2017 %>%
  mutate(total_students = rowSums(grades_2017[, 7:21]))
```

# Creating a variable using separate()

A good way to begin working with a function, to make sure you're using it right, is to consult its documentation. What's great about RStudio is that it allows you to do so within itself. Type '?separate' into the console to see its documentation. This will help you understand why we frame the *parameters* of the function the way we do.\
\

```{r, message = FALSE, tidy.opts=list(width.cutoff=60)}

grades_2017 <- grades_2017 %>% 
  separate(name, c("subject_area", NA), 
           sep = "(M|C|CM|)[0-9]", remove = FALSE)
# the regex says "(M or C or CM or <nothing>) and [any digit (from 0-9, so any digit)]"

```

There may be a few classes numbered slightly differently, meaning the 'subject area' variable will be the same as the 'name' variable in their case (since no part of their name would have matched the regular expression). Let's check, using dplyr's 'filter' function:\

```{r, message = FALSE, tidy.opts=list(width.cutoff=60)}

grades_2017 %>% filter(subject_area == name)
# only one class fails to meet numbering conventions; let's take it out
grades_2017 <- grades_2017 %>% filter(subject_area != name) 
# this command allows us to keep all the other rows.

grades_2017$subject_area <- trimws(grades_2017$subject_area)
# handy base function to trim white space

```

# Using dplyr’s group_by() and summarize()

```{r, message = FALSE, tidy.opts=list(width.cutoff=60), fig.width=4, fig.height=4}

largest_classes_subj <- grades_2017 %>% 
  group_by(subject_area) %>% 
  summarise(n = mean(total_students)) %>% 
  # gives us the sum of class enrollments 
  # grouped by subject area. 
  # replacing 'sum' with 'mean' within summarise 
  # gives mean enrollments by subject area
  arrange(desc(n)) %>%
  # arrange in descending order of n
  top_n(20)
  # select top 20 for further analysis

largest_classes_subj
```

# Using ggplot2 to plot our findings:

```{r}
largest_classes_subj %>%
  # DATASET
  ggplot(aes(x = subject_area, y = n)) +
  # ggplot() + AESTHETIC MAPPING (only axes in this case)
  geom_bar(stat = "identity") + 
  # LAYER (barplot)
  # stat = 'identity' tells ggplot not to summarize data.
  scale_x_discrete(limits = 
                     rev(largest_classes_subj$subject_area)) + 
  # SCALES (orders plot from least to most)
  coord_flip() +
  # COORDINATE SYSTEMS (flipping coordinates)
  theme_minimal()
  # changing the plot's look
  

```

# Grade distributions in Math lower-division classes:

```{r, message = FALSE, tidy.opts=list(width.cutoff=60), fig.height=6, fig.width=4}

math_lowerdivs <- which(str_detect(grades_2017$name, "MATH 3[123][AB]"))
# use str_detect() to detect the presence of the regex in a row
# wrap the which function around it to get the indices of those rows
# the regex says "MATH 3[either 1, 2 or 3][either A or B]"
math_lowerdivs <- grades_2017[math_lowerdivs, ]
# select only those rows
math_lowerdivs$name <- str_sub(math_lowerdivs$name, start = 1, end = 8) 
# use str_sub() to obtain substrings to omit whether a class is honors
math_lowerdivs <- math_lowerdivs %>% 
  gather("grade", "num_students",
         grade_Ap:grade_Cm)
# use tidyr's gather to make the wide data long
math_lowerdivs <- math_lowerdivs %>% 
  group_by(name, grade) %>% 
  summarise(n = sum(num_students))
# group by both name and grade to find 
# the number of students per grade per class

math_lowerdivs$grade <- str_sub(math_lowerdivs$grade, start = 7)
# omit "grade_" from each grade level to just get A, Ap, Am, etc.

math_lowerdivs %>% 
  ggplot(aes(x = grade, y = n)) +
  # ggplot() + AESTHETIC MAPPING (only axes)
  geom_bar(stat = "identity") +
  # LAYER (barplot)
  # stat = 'identity' tells ggplot not to summarize data.
  scale_x_discrete(limits = 
    c("Ap", "A", "Am", "Bp", "B", "Bm", "Cp", "C", "Cm")) +
  # SCALES (orders plot as desired)
  facet_wrap(~name, nrow = 3)
  # FACET (splits plot based on a variable)
```

# Some more ggplot2 with the cereal dataset:

But first, let's clean it up a bit.

```{r}
cereal <- read_csv("cereal.csv")
cereal$mfr %>% unique()
# these variables are pretty meaningless the way they're named.
# let's use str_replace_all() to rename.

# there are some -1s; not a valid value and will mess with calculations
# replace them with NA, using dplyr's na_if()
cereal <- na_if(cereal, -1)

cereal$mfr <- str_replace_all(cereal$mfr, "G", "General Mills")
cereal$mfr <- str_replace_all(cereal$mfr, "K", "Kelloggs")
cereal$mfr <- str_replace_all(cereal$mfr, "N", "Nabisco")
cereal$mfr <- str_replace_all(cereal$mfr, "P", "Post")
cereal$mfr <- str_replace_all(cereal$mfr, "Q", "Quaker Oats")
cereal$mfr <- str_replace_all(cereal$mfr, "R", "Ralston Purina")
```

# A boxplot:

```{r}
personal_palette <- c("#85D4E3", "#F4B5BD", "#E6A0C4", 
                      "#C6CDF7", "#D8A499", "#7294D4")
# colors taken from the wesanderson R color palette package

cereal %>% 
  ggplot(aes(x = mfr, y = calories, fill = mfr)) +
  geom_boxplot() +
  scale_fill_manual(values = personal_palette) +
  coord_flip() +
  theme_minimal() +
  ggtitle("Calorie content (per ounce) by manufacturer")
  # ggtitle allows you title and subtitle graphs

```

# A histogram:

```{r}
cereal %>% ggplot(aes(x = fiber, fill = mfr, color = I("black"))) + geom_histogram(bins = 10) +
  scale_fill_manual(values = personal_palette) + theme_minimal() + 
  ggtitle("Distribution of fiber across manufacturers") # don't worry about I("black")

```